Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2024-07-07 paper/paper.tex
clean up bibliography and add acknowledgments

\textcolor{red}{TODO for Fabio:
\begin{itemize}
 \item Capitalize references to algorithms, lemma etc (that is, Lemma 3.2 instead of lemma 3.2 [Lin17, Algorithm 1] instead of [Lin17, algorithm 1]), etc.
 \item Add full stops after paragraph titles.
 \item It is not great that the references occupy so much space (more than 4 pages); consider using a more condensed style. The paper needs to get strictly below 30 pages prior to submission. 
 \item Please include Acknowledgements as indicated in the comment.
\end{itemize}
}
\color{blue}
\input{intro.tex}
\input{analysis.tex}
\input{spectraldensity.tex}
\input{numresults.tex}

@@

% Acknowledge advice from David and Data/Parameters from Lin Lin, Michael Herbst?

%\clearpage
%\bibliography{bibliography.bib}
\printbibliography

@@

%\clearpage
\section{Proofs}

@@

\begin{proof}
    From \cite[theorem 3.3.2]{hogg-2013-introduction-mathematical} we know
    \begin{equation}
        \mathbb{E}^{\sfrac{p}{2}}[X] = 2 \left( \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \right)^{\sfrac{2}{p}}.
    \end{equation}
    We first treat the case $k=p=1$, for which indeed
    \begin{equation}
        2 \left( \frac{\Gamma(\frac{1+1}{2})}{\Gamma(\frac{1}{2})} \right)^2 = \frac{2}{\pi} \leq 2 = 1 + 1
    \end{equation}
    So we now assume $k + p \geq 3$. We can use a telescoping product and the property $\Gamma(z+1)/\Gamma(z) = z$ to bound
    \begin{align}
        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} 
        &= \frac{\Gamma(\frac{k}{2} + 1)}{\Gamma(\frac{k}{2})} \frac{\Gamma(\frac{k}{2} + 2)}{\Gamma(\frac{k}{2} + 1)} \cdots \frac{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor - 1)} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)} \notag \\
        &= \left(\frac{k}{2}\right) \left(\frac{k}{2} + 1\right) \cdots \left( \frac{k}{2} + \bigg\lfloor \frac{p}{2} \bigg\rfloor - 1 \right) \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)} \notag \\
        &\leq \left(\frac{k}{2} + \bigg\lfloor \frac{p}{2} \bigg\rfloor - 1\right)^{\lfloor \sfrac{p}{2} \rfloor} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)}
    \end{align}
    If $p$ is even, then $\lfloor p/2 \rfloor = p/2$ and hence
    \begin{equation}
        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} 
        \leq \left( \frac{k + p}{2} - 1 \right)^{\sfrac{p}{2}}
        \leq \left( \frac{k + p}{2}\right)^{\sfrac{p}{2}}.
    \end{equation}
    If $p$ is odd, then $\lfloor p/2 \rfloor = (p - 1)/2$, from which follows with Gautschi's inequality \cite{kershaw-1983-extensions-gautschi}
    \begin{equation}
        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})}
        \leq \left(\frac{k + p}{2} - \frac{3}{2} \right)^{ \frac{p - 1}{2}} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k + p}{2} - \frac{1}{2})}
        \leq \left(\frac{k + p}{2}\right)^{ \frac{p - 1}{2}} \left(\frac{k + p}{2}\right)^{\sfrac{1}{2}}
        = \left(\frac{k + p}{2}\right)^{ \sfrac{p}{2}}
    \end{equation}
    Hence,
    \begin{equation}
        \mathbb{E}^{\sfrac{p}{2}}[X] \leq 2 \left( \frac{k + p}{2} \right) \leq k + p.
    \end{equation}
\end{proof}

>>>>

\textcolor{red}{TODO for Fabio:
\begin{itemize}
 \item Capitalize references to algorithms, lemma etc (that is, Lemma 3.2 instead of lemma 3.2 [Lin17, Algorithm 1] instead of [Lin17, algorithm 1]), etc. \textcolor{green}{Done.}
 \item Add full stops after paragraph titles. \textcolor{green}{Done.}
 \item It is not great that the references occupy so much space (more than 4 pages); consider using a more condensed style. The paper needs to get strictly below 30 pages prior to submission. \textcolor{green}{Journal abbreviations, max 2 authors, omit superfluous fields $\to$ 3 pages.}
 \item Please include Acknowledgements as indicated in the comment. \textcolor{green}{Done.}
\end{itemize}
}
\color{blue}
\input{intro.tex}
\input{analysis.tex}
\input{spectraldensity.tex}
\input{numresults.tex}

@@

\paragraph{Acknowledgements.} The authors gratefully acknowledge the help of Lin Lin in reproducing the example in \refsec{subsec:hamiltonian}. They thank David Persson for many enlightening discussions relating the proofs in \refsec{subsec:nystrom-pp}.

\printbibliography

@@

\section{Proofs}

@@

\begin{proof}
    From \cite[Theorem 3.3.2]{hogg-2013-introduction-mathematical} we know
    \begin{equation}
        \mathbb{E}^{\sfrac{p}{2}}[X] = 2 \left( \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \right)^{\sfrac{2}{p}}.
    \end{equation}
    We first treat the case $k=p=1$, for which indeed
    \begin{equation}
        2 \left( \frac{\Gamma(\frac{1+1}{2})}{\Gamma(\frac{1}{2})} \right)^2 = \frac{2}{\pi} \leq 2 = 1 + 1
    \end{equation}
    So we now assume $k + p \geq 3$. We can use a telescoping product and the property $\Gamma(z+1)/\Gamma(z) = z$ to bound
    \begin{align}
        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} 
        &= \frac{\Gamma(\frac{k}{2} + 1)}{\Gamma(\frac{k}{2})} \frac{\Gamma(\frac{k}{2} + 2)}{\Gamma(\frac{k}{2} + 1)} \cdots \frac{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor - 1)} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)} \notag \\
        &= \left(\frac{k}{2}\right) \left(\frac{k}{2} + 1\right) \cdots \left( \frac{k}{2} + \bigg\lfloor \frac{p}{2} \bigg\rfloor - 1 \right) \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)} \notag \\
        &\leq \left(\frac{k}{2} + \bigg\lfloor \frac{p}{2} \bigg\rfloor - 1\right)^{\lfloor \sfrac{p}{2} \rfloor} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)}
    \end{align}
    If $p$ is even, then $\lfloor p/2 \rfloor = p/2$ and hence
    \begin{equation}
        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} 
        \leq \left( \frac{k + p}{2} - 1 \right)^{\sfrac{p}{2}}
        \leq \left( \frac{k + p}{2}\right)^{\sfrac{p}{2}}.
    \end{equation}
    If $p$ is odd, then $\lfloor p/2 \rfloor = (p - 1)/2$, from which follows with Gautschi's inequality \cite{kershaw-1983-extensions-gautschi}
    \begin{equation}
        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})}
        \leq \left(\frac{k + p}{2} - \frac{3}{2} \right)^{ \frac{p - 1}{2}} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k + p}{2} - \frac{1}{2})}
        \leq \left(\frac{k + p}{2}\right)^{ \frac{p - 1}{2}} \left(\frac{k + p}{2}\right)^{\sfrac{1}{2}}
        = \left(\frac{k + p}{2}\right)^{ \sfrac{p}{2}}
    \end{equation}
    Hence,
    \begin{equation}
        \mathbb{E}^{\sfrac{p}{2}}[X] \leq 2 \left( \frac{k + p}{2} \right) \leq k + p.
    \end{equation}
\end{proof}

### 2024-07-07 paper/spectraldensity.tex
clean up bibliography and add acknowledgments

It is well known that the coefficients of a Chebyshev expansion of a function can -- in exact arithmetic -- be computed by inverting the discrete cosine transform (DCT)
\begin{equation}
    g_{\sigma}(t - \cos(s_i)) = \sum_{l=0}^{m} \mu_l(t) \cos\left(\frac{\pi i l}{m} \right)
    \label{equ:discrete-cosine-transform}
\end{equation}
between the expansion coefficients $\mu_l(t),~i=0,\dots,m$ and the function evaluations at the Chebyshev nodes $s_i = \cos(\pi i / m),~i=0,\dots,m$. This can be accomplished in $\mathcal{O}(m \log(m))$ operations. \cite[algorithm 1]{lin-2017-randomized-estimation} uses a slightly different approach, which only approximates the coefficients and is noticeably slower (cf. \reftab{tab:chebyshev-timing-interpolation}).

@@

The proof specializes Bernstein's theorem for Chebyshev interpolation to the case of a Gaussian smoothing kernel \refequ{equ:smoothing-kernel}. It loosely follows the proof given in the arXiv preprint of \cite[theorem 2]{lin-2017-randomized-estimation}, but attains a slightly different dependence on $\sigma$.

\begin{proof}
    From Bernstein's theorem \cite[theorem 4.3]{trefethen-2008-gauss-quadrature} it follows that if $f$ is analytic within the Bernstein ellipse $\mathcal{E}_{\chi} \subset \mathbb{C}$ with foci $\{-1, +1\}$ and sum of semi-axes $\chi > 1$, then for any $m \in \mathbb{N}$
    \begin{equation}
        \sup_{s \in [-1, 1]} \left| f(s) - f^{(m)}(s) \right| \leq \frac{2}{\chi^m (\chi - 1)} \max_{z \in \mathcal{E}_{\chi}} |f(z)|.
        \label{equ:bernstein-bound}
    \end{equation}
    In particular, when $f = g_{\sigma}(t - \cdot)$ we observe for $z = x + iy \in \mathcal{E}_{\chi}$
    \begin{equation}
    | g_{\sigma}(t - z) | 
    = \frac{1}{\sigma \sqrt{2 \pi}} \left| e^{-\sfrac{(t - z)^2}{2\sigma^2}} \right|
    = \frac{1}{\sigma \sqrt{2 \pi}} e^{-\sfrac{(t - x)^2}{2\sigma^2}}e^{\sfrac{y^2}{2\sigma^2}}.
    %\leq \frac{1}{n \sigma \sqrt{2 \pi}} \max_{x + iy \in \mathcal{E}_{\chi}} e^{\sfrac{y^2}{2\sigma^2}} 
    \end{equation}
    Because $e^{-\sfrac{(t - x)^2}{2\sigma^2}} \leq 1$ for all $x, t \in \mathbb{R}$, and the maximum absolute value of $y$ is limited by the length of the semi-axis of the Bernstein ellipse in the direction of the imaginary axis, $(\chi - \chi^{-1}) / 2$, we upper bound
    \begin{equation}
        \max_{z \in \mathcal{E}_{\chi}} | g_{\sigma}(t - z) | 
        \leq \frac{1}{\sigma \sqrt{2 \pi}} e^{\sfrac{(\chi - \chi^{-1})^2}{8 \sigma^2}}.
    \end{equation}

@@

\Reflem{lem:chebyshev-error} shows that when $0 < \sigma \ll 1$ and if we choose $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, the error we induce with the Chebyshev expansion is at most $\mathcal{O}(\varepsilon)$. It now also allows us to bound the error we make when expanding the spectral density.

@@

\begin{table}[ht]
    \caption{We compare the runtime in milliseconds for computing the coefficients of the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ \refequ{equ:smoothing-kernel} in three different ways: as proposed in \cite[algorithm 1]{lin-2017-randomized-estimation} (\emph{oversampled FFT}), using the discrete cosine transform (\emph{DCT}) \refequ{equ:discrete-cosine-transform}, and computing a non-negative Chebyshev expansion with \refalg{alg:chebyshev-squaring} (\emph{non-negative DCT}). We use $\sigma=0.005$, $n_t=1000$ parameter values, and various degrees $m$. For stability, we average over 7 runs of the algorithms and repeat these runs 1000 times to form the mean and standard deviation which are given in the below table.}
    \label{tab:chebyshev-timing-interpolation}
   \input{tables/interpolation.tex}
\end{table}

@@

The algorithm also accomodates the two other estimators introduced in \refsec{sec:analysis}: For $n_{\mtx{\Omega}} = 0$ we recover the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}, and the complexity for each evaluation of the estimate decreases significantly. For $n_{\mtx{\Psi}} = 0$ we end up with the Nyström estimator \refequ{equ:nystrom-trace-estimator}, provided the implementation evaluates $0/0$ to $0$ in the second term on \reflin{lin:4-nystromchebyshev-nystrom-pp}. Notice that in this case, \refalg{alg:nystrom-chebyshev-pp} is equivalent to the \enquote{spectrum sweeping} method from \cite[algorithm 5]{lin-2017-randomized-estimation}, but has a significantly lower computational complexity.

@@

\begin{theorem}{Chebyshev-Nyström++ method}{chebyshev-nystrom}
    Let $\phi_{\sigma}$ be the smoothed spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$. Then the Chebyshev-Nyström++ estimate $\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, \emph{even} $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(\delta^{-1})^2)$, and smoothing parameter $0 < \sigma \ll 1$ satisfies with probability at least $1 - \delta$
    \begin{equation}
        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq \varepsilon
        \label{equ:chebyshev-nystrom-error}
        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
        %\label{equ:2-chebyshev-interpolation-error}
    \end{equation}
    for all $\varepsilon \in (0, 1)$ and $\delta \in (0, e^{-\sfrac{1}{2}}/2)$.
\end{theorem}

>>>>

It is well known that the coefficients of a Chebyshev expansion of a function can -- in exact arithmetic -- be computed by inverting the discrete cosine transform (DCT)
\begin{equation}
    g_{\sigma}(t - \cos(s_i)) = \sum_{l=0}^{m} \mu_l(t) \cos\left(\frac{\pi i l}{m} \right)
    \label{equ:discrete-cosine-transform}
\end{equation}
between the expansion coefficients $\mu_l(t),~i=0,\dots,m$ and the function evaluations at the Chebyshev nodes $s_i = \cos(\pi i / m),~i=0,\dots,m$. This can be accomplished in $\mathcal{O}(m \log(m))$ operations. \cite[Algorithm 1]{lin-2017-randomized-estimation} uses a slightly different approach, which only approximates the coefficients and is noticeably slower (cf. \reftab{tab:chebyshev-timing-interpolation}).

@@

The proof specializes Bernstein's theorem for Chebyshev interpolation to the case of a Gaussian smoothing kernel \refequ{equ:smoothing-kernel}. It loosely follows the proof given in the arXiv preprint of \cite[Theorem 2]{lin-2017-randomized-estimation}, but attains a slightly different dependence on $\sigma$.

\begin{proof}
    From Bernstein's theorem \cite[Theorem 4.3]{trefethen-2008-gauss-quadrature} it follows that if $f$ is analytic within the Bernstein ellipse $\mathcal{E}_{\chi} \subset \mathbb{C}$ with foci $\{-1, +1\}$ and sum of semi-axes $\chi > 1$, then for any $m \in \mathbb{N}$
    \begin{equation}
        \sup_{s \in [-1, 1]} \left| f(s) - f^{(m)}(s) \right| \leq \frac{2}{\chi^m (\chi - 1)} \max_{z \in \mathcal{E}_{\chi}} |f(z)|.
        \label{equ:bernstein-bound}
    \end{equation}
    In particular, when $f = g_{\sigma}(t - \cdot)$ we observe for $z = x + iy \in \mathcal{E}_{\chi}$
    \begin{equation}
    | g_{\sigma}(t - z) | 
    = \frac{1}{\sigma \sqrt{2 \pi}} \left| e^{-\sfrac{(t - z)^2}{2\sigma^2}} \right|
    = \frac{1}{\sigma \sqrt{2 \pi}} e^{-\sfrac{(t - x)^2}{2\sigma^2}}e^{\sfrac{y^2}{2\sigma^2}}.
    %\leq \frac{1}{n \sigma \sqrt{2 \pi}} \max_{x + iy \in \mathcal{E}_{\chi}} e^{\sfrac{y^2}{2\sigma^2}} 
    \end{equation}
    Because $e^{-\sfrac{(t - x)^2}{2\sigma^2}} \leq 1$ for all $x, t \in \mathbb{R}$, and the maximum absolute value of $y$ is limited by the length of the semi-axis of the Bernstein ellipse in the direction of the imaginary axis, $(\chi - \chi^{-1}) / 2$, we upper bound
    \begin{equation}
        \max_{z \in \mathcal{E}_{\chi}} | g_{\sigma}(t - z) | 
        \leq \frac{1}{\sigma \sqrt{2 \pi}} e^{\sfrac{(\chi - \chi^{-1})^2}{8 \sigma^2}}.
    \end{equation}

@@

\reflem{lem:chebyshev-error} shows that when $0 < \sigma \ll 1$ and if we choose $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, the error we induce with the Chebyshev expansion is at most $\mathcal{O}(\varepsilon)$. It now also allows us to bound the error we make when expanding the spectral density.

@@

\begin{table}[ht]
    \caption{We compare the runtime in milliseconds for computing the coefficients of the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ \refequ{equ:smoothing-kernel} in three different ways: as proposed in \cite[Algorithm 1]{lin-2017-randomized-estimation} (\emph{oversampled FFT}), using the discrete cosine transform (\emph{DCT}) \refequ{equ:discrete-cosine-transform}, and computing a non-negative Chebyshev expansion with \refalg{alg:chebyshev-squaring} (\emph{non-negative DCT}). We use $\sigma=0.005$, $n_t=1000$ parameter values, and various degrees $m$. For stability, we average over 7 runs of the algorithms and repeat these runs 1000 times to form the mean and standard deviation which are given in the below table.}
    \label{tab:chebyshev-timing-interpolation}
   \input{tables/interpolation.tex}
\end{table}

@@

The algorithm also accomodates the two other estimators introduced in \refsec{sec:analysis}: For $n_{\mtx{\Omega}} = 0$ we recover the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}, and the complexity for each evaluation of the estimate decreases significantly. For $n_{\mtx{\Psi}} = 0$ we end up with the Nyström estimator \refequ{equ:nystrom-trace-estimator}, provided the implementation evaluates $0/0$ to $0$ in the second term on \reflin{lin:4-nystromchebyshev-nystrom-pp}. Notice that in this case, \refalg{alg:nystrom-chebyshev-pp} is equivalent to the \enquote{spectrum sweeping} method from \cite[Algorithm 5]{lin-2017-randomized-estimation}, but has a significantly lower computational complexity.

@@

\begin{theorem}{Chebyshev-Nyström++ method}{chebyshev-nystrom}
    Let $\phi_{\sigma}$ be the smoothed spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$. Then the Chebyshev-Nyström++ estimate $\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, \emph{even} $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(\delta^{-1})^2)$, and smoothing parameter $0 < \sigma \ll 1$ satisfies with probability at least $1 - \delta$
    \begin{equation}
        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq \varepsilon
        \label{equ:chebyshev-nystrom-error}
        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
        %\label{equ:2-chebyshev-interpolation-error}
    \end{equation}
    for all $\varepsilon \in (0, 1)$ and $\delta \in (0, 1/2)$.
\end{theorem}
